---
title: "Plot the proportions"
author:
  - name: "Tom Smith"
    affiliation: "Cambridge Centre for Proteomics"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: | 
  Here we use the protein-level propertion object rendered in the previous notebook 
  and visualize the change in incorporation over time in the different samples.

output:
  pdf_document:
  html_notebook: default
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
---


```{r}
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
#suppressMessages(library(reshape2))
suppressMessages(library(ggplot2))
#suppressMessages(library(limma))
suppressMessages(library(MSnbase))
suppressMessages(library(camprotR))
suppressMessages(library(here))

# Standardise plotting aesthetics
source(here('R/plotting.R'))

```



```{r}
prot_proportions <- readRDS(here('results/prot_proportion.rds'))
pep_proportions_counts <- readRDS(here('results/pep_proportion_long.rds')) %>%
  filter(is.finite(ratio)) %>%
  group_by(Master.Protein.Accessions, replicate, condition) %>%
  tally()
```

Prepare proportions for plotting
```{r}
get_prot_proportions_for_plot <- function(prot_proportions){
  
  prot_proportions_for_plot <- prot_proportions %>%
    exprs() %>%
    merge(fData(prot_proportions), by='row.names') %>%
    as.data.frame() %>%
    gather(key='sample', value='proportion',
           -c(Row.names, Master.Protein.Descriptions)) %>%
    merge(pData(prot_proportions), by.x='sample', by.y='row.names') %>%
    mutate(proportion=as.numeric(proportion)) %>%
    filter(is.finite(proportion)) %>%
    mutate(type=factor(type, levels = c("3", "6", "24")))
  
  return(prot_proportions_for_plot)
}

prot_proportions_for_plot <- get_prot_proportions_for_plot(prot_proportions)

head(prot_proportions_for_plot)

```


Plot proportions
```{r}

p <- prot_proportions_for_plot %>%
  mutate(condition=recode(condition, 'untrt'='Control')) %>%
  ggplot(aes(proportion, colour=condition, linetype=replicate)) +
  geom_density() +
  theme(axis.text=element_text(size=10)) +
  facet_wrap(~as.numeric(replicate)) +
  scale_colour_manual(values=c(get_cat_palette(3), 'black'), name='') +
  xlab('Proportion nascent protein') +
  facet_grid(.~type) +
  scale_linetype_discrete(name='Replicate') +
  ylab('Density') +
  scale_x_continuous(limits=c(0,1), breaks=seq(0,1,0.2))
      
    
print(p)
ggsave(here('notebooks/plots/proportion_nascent_dist.png'))
ggsave(here('notebooks/plots/proportion_nascent_dist.pdf'))

```
Compare proportions between A549 and MCF10A. The correlation is not great --> A549 has faster protein turnover than MCF10A, as would be expected from a tumoral line. 



```{r}
compare_proportions_conditions <- prot_proportions_for_plot %>%
      select(-sample) %>%
      pivot_wider(values_from=proportion, names_from=condition)

saveRDS(compare_proportions_conditions, here('results/compare_proportions_conditions.rds'))

compare_proportions_type <- prot_proportions_for_plot %>%
      select(-sample) %>%
      pivot_wider(values_from=proportion, names_from=type)

saveRDS(compare_proportions_type, here('results/compare_proportions_type.rds'))
  



```




Plot correlation of proportions between DMSO & Tg. Median correlation here and overall lower nascent in Tg, especially for RNA-bound, as expected.
```{r, fig.height=7, fig.width=7}
condition_proportion_cor_per_rep <- compare_proportions_conditions %>%
  filter(is.finite(GC7), is.finite(untrt)) %>%
  group_by(type, replicate) %>%
  summarise(cor=cor(GC7, untrt, use='complete', method='spearman'))
  
compare_proportions_conditions %>%
  ggplot(aes(y=100*GC7, 100*untrt)) +
  xlim(0, 100) +
  ylim(0, 100) +
  geom_abline(slope=1, linetype=2, colour='grey') +
  geom_point(size=0.5, alpha=0.5) +
  geom_text(data=condition_proportion_cor_per_rep, inherit.aes=FALSE,
            aes(label=paste0('rho=', sprintf("%.2f", cor))), x=0, y=100, hjust=0, size=3) +
  facet_grid(replicate~type) +
  ylab('Proportion nascent in GC7') +
  xlab('Proportion nascent in untrt') +
  theme(text=element_text(size=15))
```   
OK, great, so there is a good correlation between the conditions, but at 3hr there seem to be more proteins off the diagonal (higher or lower translation with GC7 inhibition). I suspect this is just less accurate quantification. Later, we'll see if these are consistent differences.



